### 2018-09-04 doc/intro.tex

Species extinction risk is strongly linked to geographic range size.

A species' geographic range size changes over time.

How does extinction risk change over a species duration? What effect does geographic range have on this?

We want to know if including information about past geographic range size improves our inference of species extinction risk.

A lot of work on species ``rise and fall'' dynamics has focused on the change in species geographic range over timeand what the shape of the average trajectory is CITATION. Issues that aren't addressed, like non-random drop out. Species can go extinct at any point in their ``lifetime'', so we might not observe a ``pure'' change to geographic range. We have to take this into account. The models that have been used do not do that which can totally mess-up analysis of geographic ranges over time. They need a joint-model, but that hasn't penetrated the field yet.

Additionally, lack of mechanism to explain a pure ``rise-fall''. Importantly, ``rise and fall'' is expected from a Markov process that has only a single absorbing boundary; at some point it will go back to 0.

To demonstrate that species geographic range size is not a complete Markov process, it would be necessary to show that extinction risk depends not just on the current geographic range size, but the history of its geographic range size: has range size declined or increased between earlier and now?

We instead are focusing on how geographic range effects extinction risk of a species. We are not modeling geographic range through time, but instead modeling how species extinction is a function of geographic range over time.

One of the great promises of paleobiology is that by studying the past we can better predict the future.  This promise is particularly pertinent given suggestions that risk assessments for some modern species could be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.  Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?

@@

how predictable is extinction? how much does the past improve our predictions? how much does the past improve our predictions?

we know that at almost all times, a species with geater than average geographic range is expected to have a lower extinction risk than a closely related species with a lower than average geographic range. what we don't know is if expanding or decreasing in geographic range over million-year time scales has any additional predictive information about a species' risk of going extinct.

>>>>

How predictable is extinction? Forecasting species survival at million-year timescales.

One of the great promises of paleobiology is that by studying the past we can better predict the future.  This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.  Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?

@@

Species global geographic range changes over time due to both local and global processes.

An old assumption in paleobiology was that species quickly obtained their maximum global geographic range and then maintained that range till their decline eventual extinction. Observation of species distribution records does not support this assumption. Instead, changes to a species' global geographic range is currently theorized to follow one of two general patterns: hat or hump PERFECT TIME FOR A FIGURE. Both are symmetric patterns but have different shapes. The hat pattern is characterized by an increase in geographic range for the first half a species' duration followed by a brief peak in geographic range, and the a decrease in geographic range for the second half of that species' duration. The hump pattern is instead characterized by a brief period of increasing species geographic range, an extended period of that species' duration at its maximum geographic range, and finally a quick decline and extinction of the species.

Measuring this pattern is extremely difficult for multiple reasons. Species are not observed continuously through time, species geographic ranges are based on what samples are avaliable, and species are harder to observe at the tails of their duration because they are rare (small geographic range, small population sizes).

Additionally, neither of these patterns are necessarily different from a Markovian process with a lower bound. A random walk beginning at 0 will most likely be unimodal with a period of increase followed by a period of decrease and then eventually reaching 0. PERFECT PLACE FOR A SIMULATION.

### 2018-09-04 doc/methods.tex

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\). These regression coefficients are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept.

@@

To complete the generative model, we need to assign final priors to the ``top level'' parameters
\begin{equation}
  \begin{aligned}
    \mu_{c} &\sim 
      \begin{cases}
        N(0, 10) & \text{if } c = 1 \\
        N(0, 3) & \text{if } c > 1 \\
      \end{cases}
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{a} &= diag(\tau_{a}) \Omega_{a} diag(\tau_{a}) \\
    \Sigma_{\delta} &= diag(\tau_{\delta}) \Omega_{\delta} diag(\tau_{\delta}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{a} &\sim C^{+}(5) \\
    \tau_{\delta} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{a} &\sim LKJ(1) \\
    \Omega_{\delta} &\sim LKJ(1). \\
  \end{aligned}
  \label{eq:priors}
\end{equation}
The decomposition of the covariance matrices (e.g \(\Sigma_{B}\)) allows priors the variance and correlation aspects of covariance to be defined separately. These priors are also slightly more interpretable than a, for example, inverse-Wishart distributed prior for the covariance matrices. This decomposition is recommended by CITATION.

@@

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.99; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION. An example line of R code to fit this mode given a data.frame of all necessary data called ``data'' could be written as:
\begin{verbatim}
  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
                       (1 + range + range_diff | mybin/phylum) + 
                       (1 | age/phylum), 
             prior = normal(0, 3, autoscale = FALSE), 
             prior_intercept = normal(0, 10, autoscale = FALSE), 
             prior_aux = cauchy(0, 5, autoscale = FALSE), 
             data = data, 
             adapt_delta = 0.99, 
             thin = 4)
\end{verbatim}

@@

We considered four variants of this model: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect).

To determine how many non-nested varying-intercept terms were possible to include without overly biasing out models' parameter estimates, the four variant models described above were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior CITATION. WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance. Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for a more nuanced understanding the actual amount of difference between the models. We selected the model with the lowest WAIC and LOOIC, which happened to be are most parameter rich model.

Model adequacy was measured using the area under the receiver operating characteristic curve (AUC). This measure is commonly used in classification problems like this one as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION. This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC was calculated for the dataset as a whole, and for each of time bins. These values are measures of in-sample predictive accuracy.

\subsection{Prediction given class imbalance}

@@

\subsection{Cross-validation}

In addition to model selection and adequacy, we are also interested in understanding how well our model predicts species extinction given new,future data. We estimated out-of-sample predictive error using cross-validation. For our dataset, expected out-of-sample AUC was estimated using five-fold cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

For each posterior draw of the cross-validation predictions, the optimal cut-point was calculated as described above. Given these class assignments based on estimated optimal cut-point, the AUC value of the out-of-sample estimates was calculated. AUC values are calculated for the entire fold and for each of the individual time points represented within that fold.

>>>>

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\). These regression coefficients are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept. \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.

@@

To complete the generative model, we need to assign final priors to the ``top level'' parameters. In general we favored weakly informative priors which help regularize our estimates. In the case of a regression coefficient, this means a Normal distribution with mean 0 and a standard deviation of 3. For our scale parameters (e.g. standard deviations), we used half-Cauchy distributed priors with heavy tails but the majority of probability density near 0.

Our top-level intercept was given a more diffuse prior than our regression coefficients, which reflects our greater degree of uncertainty about its value. Our top-level regression coefficient for the effect of geographic range was given an informative prior reflecting the overwhelming amount of evidence that species with a larger than average geographic range have a lower risk of extinction than species with an average or less than average geographic range. In the context of this analysis, this means that we are again using a weakly informative prior but instead of centering the density around -1 (i.e. larger than average geographic range decreases extinction risk).

Instead of assigning a prior distribution for each of the covariance matrices in the model, we instead decomposed the covariance matrices (e.g. \(\Sigma_{B}\)) which allows us to assign independent priors for the scale and correlation aspects of covariance. The scale parameters were assigned half-Cauchy priors as described above in the context of all other scale parameters. The correlation matrices were assigned LKJ priors each with shape parameter set to 1. This choice of shape parameter produces a uniform distribution over possible correlation matrices. These priors are also slightly more interpretable than other common prior distributions for covariance matrices such as the inverse-Wishart distribution. This approach to assigning priors to a covariance matrix is recommended by CITATION.

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(0, 10) & \text{if } d = 1 \\
        N(-1, 3) & \text{if } d = 2 \\
        N(0, 3) & \text{if } d > 2 \\
      \end{cases}
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{a} &= diag(\tau_{a}) \Omega_{a} diag(\tau_{a}) \\
    \Sigma_{\delta} &= diag(\tau_{\delta}) \Omega_{\delta} diag(\tau_{\delta}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{a} &\sim C^{+}(5) \\
    \tau_{\delta} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{a} &\sim LKJ(1) \\
    \Omega_{\delta} &\sim LKJ(1). \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

@@

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION. An example line of R code to fit this mode given a data.frame of all necessary data called ``data'' could be written as:
\begin{verbatim}
  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
                       (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                       (1 | age/phylum), 
             prior = normal(c(-1, 0, 0, 0), rep(3, 4), autoscale = FALSE), 
             prior_intercept = normal(0, 10, autoscale = FALSE), 
             prior_aux = cauchy(0, 5, autoscale = FALSE), 
             data = data, 
             adapt_delta = 0.9999, 
             thin = 4)
\end{verbatim}

@@

We considered four variants of the model described above: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect). The second and fourth models modify the number of columns in \(X\) by removing two of the covariates. The third and fourth model simplify the first model to just varying intercepts without varying slopes.

The four variant models were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior CITATION. WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance. Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for a more nuanced understanding the actual amount of difference between the models. We selected the model with the lowest WAIC and LOOIC, which happened to be are most parameter rich model.

Model adequacy was measured using the area under the receiver operating characteristic curve (AUC). This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION. This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC was calculated for the dataset as a whole, and for each of time bins. This measure was used in two contexts: measuring in-sample and out-of-sample predictive performance. ``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. ``Out-of-sample'' is defined below.

We are also interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data). To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

@@
